Optimizing Abstract Abstract Machines

### 2012-11-26 paper/paper.tex
Drop .pdf from includegraphics.
Fixed makefile to build final visualizations, renamed memory to speed for consistency.
Implemented some of Mitch's comments.

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
\mbox{ where }\maddr \notin\msto
\\
\mathit{bind}_\mlab^\mcntr(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
\mbox{ where }\maddr \notin\msto
\end{align*}
The resulting relation is non-deterministic in its choice of
addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming, this relation defines
a deterministic machine.  (The relation is really a function.)

@@

\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
                                           \msto\sqcup[\maddr \mapsto
                                             \{ \mval\}],
                                           \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= x\mcntr' \\
%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
\end{align*}
where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto
\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup \msto'(\maddr)$.
The \(\rfloor\cdot\lfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

To describe this process, we'll refactor the transition relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The refactored relation, $\widehat{\machstep}$, becomes:
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c, \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all subsequent stores.

@@

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially.
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.

If we specialize to 0CFA, the evaluation rules become:

@@

\begin{align*}
%ext(\msto, \maddr, s) &= \msto \sqcup [\maddr \mapsto s] \\
\ev{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev{\mexpi{0},\menv,\ext{\msto}{\mlab}{\{\mkont\}},\kar{\mexpi{1},\mlab}}
\\
\co{\kar{\mexp,\menv,\mlab},\mval,\msto} & \machstep
\ev{\mexp,\menv,\ext{\msto}{\mlab^f}{\{\mval\}},\kfn{\mlab^f,\mlab}}
\\
\co{\kfn{\mlab^f,\mlab},\mval,\msto} & \machstep
\ap[_\mlab]{\mvalx{u},\mlab^a,\mkont,\ext{\msto}{\mlab^a}{\{\mval\}}}
\\
\text{ where } &\mkont \in \msto(\maddr), \mvalx{u} \in \msto(\maddr^f)
\\
\ap[_\mlab]{\clos{\mvar,\mexp,\menv},\mlab^a, \msto,\mkont} & \machstep
\ev{\mexp,\menv[\mvar\mapsto\mvar],\ext{\msto}{\mvar}{\msto(\mlab^a)},\mkont}
\\
\ap[_\mlab]{\mop,\mlab^a,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\\
\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\mlab^a)\}
\end{align*}

@@

Tracing the execution of the analysis reveals an immediate shortcoming:
there is a high degree of branching and merging in the exploration.
%
Surveying this branching has no benefit for precision.
%
For example, in a function application, {\tt (f x y)},
where {\tt f}, {\tt x} and {\tt y} each have several values
each argument evaluation induces two-way branching, only to be ultimately joined back together in their respective
application positions. 
%
Transition patterns of this shape litter the state-graph:
%
\begin{center}
\includegraphics[scale=0.3]{fanout.pdf}
\end{center}
To avoid the spurious forking and joining, we {\it delay} the non-determinism
until and unless it is needed in {\it strict contexts} (such as the guard for an
{\tt if} a called procedure, or a numerical primitive application). 
%
Doing so collapses these forks and joins into a linear sequence of states:
\begin{center}
\includegraphics[scale=0.3]{lazy.pdf}
\end{center}

In case it's unclear, this shift does not change the concrete semantics of the
language to be lazy.  Rather, it abstracts over transitions that the original
non-deterministic semantics steps through. 
%
We say the abstraction is \emph{lazy} because it delays dereferencing an
address until its contents are \emph{needed} as values in the semantics. It
does not change the execution order that leads to the values that are stored in
the address.

@@

Quotation is special because it cannot introduce function values,
which is important to enhancing our technique's soundness to
conceptual complexity ratio. If we join two values of different type
together, we don't get ``anything,'' which has complex meaning in
higher-order languages (Shivers' escape semantics) and is overly
approximate. We instead get ``any quotable value,'' which has much
simpler semantics.

@@

@@

We have implemented, optimized, and evaluated an analysis framework
supporting higher-order functions, state, first-class control,
compound data, and a large number of primitive kinds of data and
operations such as floating point, complex, and exact rational
arithmetic.  The analysis is evaluated against a suite of benchmarks
drawn from the literature.
%
For each benchmark, we collect analysis times, peak memory usage, and
the rate of states-per-second explored by the analysis for each of the
optimizations discussed in section~\ref{sec:opt}, cumulatively
applied.  The analysis is stopped after consuming 30 minutes of time
or 1 gigabyte of space.  When presenting \emph{relative} numbers, we
use the timeout limits as a lower bound on the actual time required,
thus giving a conservative estimate of improvements.

@@

Many benchmarks cause the baseline analyzer to take longer than 30
minutes or to consume more 1 gigabyte of memory, at which point the
analysis is stopped.  This is the case for the largest benchmark
program, which is 3,500 lines of code and takes under a minute in the
most optimized analyzer.  For those benchmarks that did complete on
the baseline, the optimized analyzer outperformed the baseline by a
factor of two to three orders of magnitude.
% safer to not give exact factors
% 475 to 4,382.

@@

\item {\bf nbody}: implements the Greengard multipole algorithm for
  computing gravitational forces on point masses distributed uniformly
  in a cube (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).

@@

The analysis considered here computes results similar Earl, et al.'s
0-CFA implementation~\cite{dvanhorn:Earl2012Introspective}, which
times out on the \Church{} benchmark because it does not widen the
store as described for our baseline evaluator.  So even though it
offers a fair point of comparison, a more thorough evaluation is
probably uninformative as the other benchmarks are likely to timeout
as well (and it would require significant effort to extend their
implementation with the features needed to analyze our benchmark
suite).  That implementation is evaluated against much smaller
benchmarks: the largest program is 30 lines.

@@

Constraint-based program analyses
(e.g.~\cite{dvanhorn:nielson-nielson-popl97,dvanhorn:wright-jagannathan-toplas98,dvanhorn:Meunier2006Modular})
typically compute sets of abstract values for each program point.
These values approximate values arising at run-time for each
program point.  Value sets are computed as the least solution to a set
of (inclusion or equality) constraints.  The
constraints must be designed and proved as a sound approximation of
the semantics.  Efficient implementations of these kinds of analyses
often take the form of worklist-based graph algorithms for constraint
solving, and are thus quite different from the interpreter
implementation.  The approach thus requires effort in constraint
system design and implementation, and the resulting system require
verification effort to prove the constraint system is sound and that
the implementation is correct.

>>>>

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
\mbox{ where }\maddr \notin\msto
\\
\mathit{bind}_\mlab^\mcntr(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
\mbox{ where }\maddr \notin\msto
\end{align*}
The resulting relation is non-deterministic in its choice of
addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming and fix an allocation
scheme, this relation defines a deterministic machine (the relation is really a function).

@@

\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
                                           \msto\sqcup[\maddr \mapsto
                                             \{ \mval\}],
                                           \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= x\mcntr' \\
%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
\end{align*}
where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto
\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup \msto'(\maddr)$.
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

To describe this process, we'll refactor the transition relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The refactored relation, $\widehat{\machstep}$, becomes:
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all stores after stepping.

@@

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially. In our example language, we have only
unary functions, so application forms always have two positions: the
function and the argument. Thus for every application labeled $\mlab$,
we introduce two new labels, $\mlab^f$ and $\mlab^a$.
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.

In $k$-CFA, the evaluation rules become:

@@

\begin{align*}
%ext(\msto, \maddr, s) &= \msto \sqcup [\maddr \mapsto s] \\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi{0},\menv,\ext{\msto}{\mlab}{\{\mkont\}},\kar{\mexpi{1},\menv,\mlab,\mcntr}}
\\
\co{\kar{\mexp,\menv,\mlab,\mcntr},\mval,\msto} & \machstep
\ev[^\mcntr]{\mexp,\menv,\ext{\msto}{\mlab^f\mcntr}{\{\mval\}},\kfn{\mlab^f,\mlab,\mcntr}}
\\
\co{\kfn{\mlab^f,\mlab,\mcntr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mvalx{u},\mlab^a,\mkont,\ext{\msto}{\mlab^a\mcntr}{\{\mval\}}}
\\
\text{ where } &\mkont \in \msto(\mlab\mcntr), \mvalx{u} \in \msto(\mlab^f\mcntr)
\\
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mlab^a, \msto,\mkont} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv[\mvar\mapsto\mvar\mcntr'],\ext{\msto}{\mvar\mcntr'}{\msto(\mlab^a\mcntr)},\mkont}
\\
\text{ where }& \mcntr' = \lfloor\mlab\mcntr\rfloor_k
\\
\ap[^\mcntr_\mlab]{\mop,\mlab^a,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\\
\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\mlab^a\mcntr)\}
\end{align*}

@@

Tracing the execution of the analysis reveals an immediate shortcoming:
there is a high degree of branching and merging in the exploration.
%
Surveying this branching has no benefit for precision.
%
For example, in a function application, {\tt (f x y)},
where {\tt f}, {\tt x} and {\tt y} each have several values
each argument evaluation induces two-way branching, only to be ultimately joined back together in their respective
application positions. 
%
Transition patterns of this shape litter the state-graph:
%
\begin{center}
\includegraphics[scale=0.3]{fanout}
\end{center}
To avoid the spurious forking and joining, we {\it delay} the non-determinism
until and unless it is needed in {\it strict contexts} (such as the guard for an
{\tt if} a called procedure, or a numerical primitive application). 
%
Doing so collapses these forks and joins into a linear sequence of states:
\begin{center}
\includegraphics[scale=0.3]{lazy}
\end{center}

This shift does not change the concrete semantics of the language to
be lazy.  Rather, it abstracts over transitions that the original
non-deterministic semantics steps through.
%
We say the abstraction is \emph{lazy} because it delays dereferencing an
address until its contents are \emph{needed} as values in the semantics. It
does not change the execution order that leads to the values that are stored in
the address.

@@

Quotation is special because it cannot introduce function values,
which is important for a simple semantics. If we join two values of
different type together, we don't get ``anything,'' which has complex
meaning in higher-order languages (Shivers' escape semantics) and is
overly approximate. We instead get ``any quotable value,'' which has
much simpler semantics.

@@

This abstraction was so critical to timely convergence that all of our evaluation uses it.

@@

We have implemented, optimized, and evaluated an analysis framework
supporting higher-order functions, state, first-class control,
compound data, and a large number of primitive kinds of data and
operations such as floating point, complex, and exact rational
arithmetic.  The analysis is evaluated against a suite of Scheme benchmarks
drawn from the literature.
%
For each benchmark, we collect analysis times, peak memory usage, and
the rate of states-per-second explored by the analysis for each of the
optimizations discussed in section~\ref{sec:opt}, cumulatively
applied.  The analysis is stopped after consuming 30 minutes of time
or 1 gigabyte of space.  When presenting \emph{relative} numbers, we
use the timeout limits as a lower bound on the actual time required,
thus giving a conservative estimate of improvements.

@@

Many benchmarks cause the baseline analyzer to take longer than 30
minutes or to consume more 1 gigabyte of memory, at which point the
analysis is stopped.  This is the case for the largest benchmark
program, {\bf nucleic}, which is 3,500 lines of code and takes under a minute in the
most optimized analyzer.  For those benchmarks that did complete on
the baseline, the optimized analyzer outperformed the baseline by a
factor of two to three orders of magnitude.
% safer to not give exact factors
% 475 to 4,382.

@@

\item {\bf nbody}: implementation~\cite{ianjohnson:nbody87} of the
  Greengard multipole algorithm for computing gravitational forces on
  point masses distributed uniformly in a cube (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).

@@

The analysis considered here computes results similar to Earl, et al.'s
0-CFA implementation~\cite{dvanhorn:Earl2012Introspective}, which
times out on the \Church{} benchmark because it does not widen the
store as described for our baseline evaluator.  So even though it
offers a fair point of comparison, a more thorough evaluation is
probably uninformative as the other benchmarks are likely to timeout
as well (and it would require significant effort to extend their
implementation with the features needed to analyze our benchmark
suite).  That implementation is evaluated against much smaller
benchmarks: the largest program is 30 lines.

@@

Constraint-based program analyses
(e.g.~\cite{dvanhorn:nielson-nielson-popl97,dvanhorn:wright-jagannathan-toplas98,dvanhorn:Meunier2006Modular,dvanhorn:steckler-wand-toplas97})
typically compute sets of abstract values for each program point.
These values approximate values arising at run-time for each
program point.  Value sets are computed as the least solution to a set
of (inclusion or equality) constraints.  The
constraints must be designed and proved as a sound approximation of
the semantics.  Efficient implementations of these kinds of analyses
often take the form of worklist-based graph algorithms for constraint
solving, and are thus quite different from the interpreter
implementation.  The approach thus requires effort in constraint
system design and implementation, and the resulting system require
verification effort to prove the constraint system is sound and that
the implementation is correct.
